# Logistic regression

## Data
The dataset has been constructed by joining two student alchol comsuption datasets which were downloaded from https://archive.ics.uci.edu/ml/datasets/Student+Performance. The two datasets contain the same variables and the students are partially overlapping. The data was joined by using following columns as surrogate identifiers for students:

* *school, sex, age, address, famsize, Pstatus, Medu, Fedu, Mjob, Fjob, reason, nursery, internet*

The variables not used for joining have been combined by averaging for numeric columns and by taking the first answer for non-numeric columns. Two new variables have been define:

* *alc_use* is the average of *Dalc* and *Walc*
* *high_use* is TRUE if *alc_use* is hihger than 2 and FALSE otherwise

Read in the dataset.

```{r data_reading_e3}
alc <- read.csv("~/git/IODS-project/data/alc.csv")
```

Variables in the data set:
```{r data_check_e3}
str(alc)
```

## Predictors of high alchol consumption

### Hypothesis
I choose the following 4 variables for studying their relationship to high/low alcohol consumption:

* *goout*  - going out with friends (numeric: from 1 - very low to 5 - very high)
* *sex* - student's sex (binary: 'F' - female or 'M' - male)
* *studytime* - weekly study time (numeric: 1 - <2 hours, 2 - 2 to 5 hours, 3 - 5 to 10 hours, or 4 - >10 hours)
* *romantic* - with a romantic relationship (binary: yes or no) 

I hypotize that high amount of "going out with friends" would be associated with a higer alcohol consumption since alcohol is often consumed in social situations. I think male gender could be associated to higher alcohol consumption just because males can tolerate more alcohol. I hypotize that higher "weekly study time" is associated to low alcohol compsuption: students with high study times are focused on school and don't have as much time to drink alcohol. I also hypotize that individuals "with a romatic relationship" will consume less aclhol since they spend more time with their partners than friens and alchol is more often spend with friends.

### Data exploration
Explore the variables of interest in regard to alchohol use

```{r exploratory_plots_3, fig.height=12, fig.width=12, warning=F, message=F}
library(tidyr)
library(dplyr)
library(ggplot2)

# Explore the mean 'goout' and 'studytime' with respect to high/low alcohol use
alc %>% group_by(high_use) %>% summarise(count = n(), mean_goout=mean(goout),
                                         mean_studytime=mean(studytime))

# Explore high/low alcohol use stratified by 'sex'
alc %>% group_by(high_use, sex) %>% summarise(count = n())

# Explore high/low alcohol use stratified by 'romantic'
alc %>% group_by(high_use, romantic) %>% summarise(count = n())

# Draw barplots of variables of interest
###########################################

g_goout <- ggplot(alc, aes(x = goout, fill=high_use)) +
  geom_bar() + xlab("Going out with friends") +
  ggtitle("Going out with friends from 1 (very low) to 5 (very high) by alcohol use")

g_studytime <- ggplot(alc, aes(x = studytime, fill=high_use)) +
  geom_bar() + xlab("Weekly study time") +
  ggtitle("Weekly study time [1 (<2 hours), 2 (2 to 5 hours), 3 (5 to 10 hours), or 4 (>10 hours)] by alchol use")

g_sex <- ggplot(alc, aes(x = sex, fill=high_use)) +
  geom_bar() + 
  ggtitle("Sex by alcohol use")

g_romantic <- ggplot(alc, aes(x = romantic, fill=high_use)) +
  geom_bar() + 
  ggtitle("With a romantic relationship (yes/no) by alcohol use")

# Arrange the plots into a grid
library("gridExtra")
grid.arrange(g_goout, g_studytime, g_sex, g_romantic, ncol=2, nrow=2)


```

Based on the plots my assumptions seem to be somewhat corret:

* High alchol use is associated with going out with friends
* Low alchol use is asscoated with high weekly study time
* High alchol use is more common in men compared to women
* Low alchol use is more common in individuals in romatic relationships

### Fitting a logistic regression model

Fit a logistic regression model using *high_use* as the targe variable and *goout*, *studytime*, *sex* and *romantic* as the explanatory variables.

```{r logistic_regression_3}
# Fit the logistic regression model
m <- glm(high_use ~ goout + studytime + sex + romantic, data = alc, family = "binomial")
```

Summary of the fitted logistics regression model shows that *goout*, *studytime* and *sex* are statistically significantly associated to alchol comsumption. High alcohol comsumption is associated to high *goout* and male gender and low alcohol comsumption is associated to high *studytime*.

```{r logistic_regression_summary_3}
# Summary of the model
summary(m)
```

Coefficients of the model as odds ratios and their confidence intervals:
```{r logistic_regression_or_3, message=F}
# Calculate the odds ratios and confidence intervals of the coefficients
or <- coef(m) %>% exp
ci <- confint(m) %>% exp
# Print out the odds ratios and confidence intervals
cbind(or, ci)
```
From the odds ratios we can see that one unit increase in *goout* is associated with 2.1 times higher likelihood of high alchohol comsumption and one unit increase in *studytime* is associated with 0.6 times lower likelihood of high alcohol comsumption. Male gender is associated with 1.9 higer likelihood of high alcohol comsumption comapred to female gender. Being in a romatic relationship is not significantly associated to high/low alcohol comsuption in this model since the confidence intervals include 1.

My previously stated hypothesis seem to be verified by this model except that *romantic* is not associated to high or low alcohol comsumption in this model.

### Performance of the model
Fit a logistic model with the explanatory variables that were statistically significantly associated to high or low alchohol consumption:

```{r}
# Fit the logistic regression model
m <- glm(high_use ~ goout + studytime + sex, data = alc, family = "binomial")
```

Prediction performance of the model

```{r}
# Calcualte the predicted probabilities of high alcohol comspumtion
probability <- predict(m, type="response")
alc <- mutate(alc, probability=probability)
# Predict the high alcohol use with the probabilities
alc <- mutate(alc, prediction=probability > 0.5)
# Cross-tabulate the actual class and the predicted class
table(high_use = alc$high_use, prediction = alc$prediction)

```
The model seems to be quite good at predicting low alcohol use but performes less well in predicting high alcohol use.

Visualize the actual class, the predicted probabilities and the predicted class.

```{r}
# Initialize a plot of 'high_use' versus 'probability' in 'alc'
g <- ggplot(alc, aes(x = probability, y = high_use, col=prediction))

# define the geom as points and draw the plot
g + geom_point()

```

Calculate the total proportion of mis-classified individuals using the regression model and with a simple guessing strategy where everyone is classified to be in the most prevalent class low use of alcohol.
```{r}
# Define a loss function (mean prediction error)
loss_func <- function(class, prob) {
  n_wrong <- abs(class - prob) > 0.5
  mean(n_wrong)
}

# Call loss_func to compute the average number of wrong predictions in the (training) data
loss_func(class = alc$high_use, prob = alc$probability)

# Compare the results to guessing that everybody belongs to the class low use of alcohol
loss_func(class = alc$high_use, prob = 0)
```
Using the regression model 24.6% of the individuals are mis-classified compared to 29.8 % of mis-classified individuals when simply guessing everybody belongs to the low use of alcohol class. The model seems to provide some improvement to the simple guess of the most prevalent class.

### Cross-validation of the model
Performe 10-fold cross-validation of the model to estiamte the performance of the model on unseen data. The performance of the model is measured with proportion of mis-classified individuals. The mean prediction error in the test set:

```{r}
library(boot)
# 10-fold cross-validation
cv <- cv.glm(data = alc, cost = loss_func, glmfit = m, K = 10)
# Mean prediction error
cv$delta[1]
```
The mean prediction error in the test set is 0.25 which is better than the performance of the model introduced in the DataCamp exercises which had a mean prediction error of 0.26 in the test set.


### Models with different number of predictors
Construct models with different number of predictors and calculate the test set and training set prediction errors.

```{r}
# All the possible predictors
predictors <- c('school', 'sex', 'age', 'address', 'famsize', 'Pstatus', 'Medu',
                'Fedu', 'Mjob', 'Fjob', 'reason', 'nursery', 'internet', 'guardian',
                'traveltime', 'studytime', 'failures', 'schoolsup', 'famsup', 'paid',
                'activities', 'higher', 'romantic', 'famrel', 'freetime', 'goout',
                'health', 'absences', 'G1', 'G2', 'G3')


# Fit several models and record the test and traingin errors
# 1) Use all of the predictors.
# 2) Drop one predictor and fit a new model.
# 3) Continue until only one predictor is left in the model.


# Fit the models and calculate the erros
test_error <- numeric(length(predictors))
training_error <- numeric(length(predictors))

for(i in length(predictors):1) {
  model_formula <- paste0("high_use ~ ", paste(predictors[1:i], collapse = " + "))
  glmfit <- glm(model_formula, data = alc, family = "binomial")

  # 10-fold cross-validation
  cv <- cv.glm(data = alc, cost = loss_func, glmfit = m, K = 10)
  # Mean prediction error
  test_error[i] <- cv$delta[1]
  # Training error
  training_error[i] <- 
    loss_func(alc$high_use, predict(glmfit,type="response"))
}

# Construct a table of prediction errors for plotting
data_error <- rbind(data.frame(n_predictors=1:length(predictors),
                               prediction_error=test_error,
                               type = "test error"),
                    data.frame(n_predictors=1:length(predictors),
                               prediction_error=training_error,
                               type = "training error"))

                    
# Plot the test and training errors vs. number of predictors in the model
g <- ggplot(data_error, aes(x = n_predictors, y = prediction_error, col=type))

# define the geom as points and draw the plot
g + geom_point()

```

