# Regression and model validation

## Data
Read in the dataset. The dataset contains 7 variables:

* *gender*: Gender: M (Male), F (Female)
* *age*: Age (in years) derived from the date of birth
* *attitude*: Global attitude toward statistics
* *deep*: Deep approach
* *surf*: Surface approach
* *stra*: Strategic approach
* *points*: Exam points

```{r data_reading}
learning2014 <- read.csv("~/git/IODS-project/data/learning2014.csv")
```

Check that the data was read correctly. Print out the first few rows and check the type of the columns:
```{r data_check}
head(learning2014)
str(learning2014)
```


Explore the data by plotting
```{r exploratory_plots, fig.height=12, fig.width=12, warning=F, message=F}
library(GGally)
library(ggplot2)

# Define the plot.
p <- ggpairs(learning2014, mapping = aes(col = gender, alpha=0.3), 
             lower = list(combo = wrap("facethist", bins = 20)), legend=1)

# Draw the plot
p

```

There are almost twice as many women in the dataset compared to men. The distribution of *age* is skewed towards higer values. The distribution of *attitutde*, *deep*, *surf*, *stra* and *points* seems to be more normally distributed. The variable with the highest correlation with *points* is *attitude*. 

```{r summaries_of_the_variables}
summary(learning2014$gender)
summary(learning2014$age)
summary(learning2014$attitude)
summary(learning2014$deep)
summary(learning2014$surf)
summary(learning2014$stra)
summary(learning2014$points)

```


##   Regression analysis
Fit a regression model using **points** as the outcome variable and  **age**, **gender** and **attitude** as explanatory variables:

```{r}
# Fit the linear regression model
lm_model <- lm(points ~ age + gender + attitude, data=learning2014)

# Sumamry of the model
summary(lm_model)

```
*Gender* and *age* are not significantly associated to the points so remove them from the model.
Fit the model again using only *attitude* as the explanatory variable.

```{r}
lm_model_2 <- lm(points ~ attitude, data=learning2014)
summary(lm_model_2)
```

Explanatory variable *attitude* is statistically significantly associated to the outcome variable *points* with a p-value of 4.12e-09. The estimate of the regression coefficient is 0.35 meaning a one unit increase in *attitude* is on average associated to a 1.42 unit increase in *points*. The R² of the model is 0.19 which means that the variation of the explanatory variable *attitude* explains 19% of the variation of the outcome variable *points*.

## Diagnostics of the regression model
Plot the residuals versus the fitted values 
```{r residuals_fitted_values_plot}
plot(lm_model_2, which=1)
```

Plot the normal QQ-plot:
```{r normal_qq_plot}
plot(lm_model_2, which=2)
```

Plot the residuals versus leverage plot:
```{r residuals_leverage_plot}
plot(lm_model_2, which=5)
```


